Psychological Needs During Intergroup Contact

### 2022-01-17 manuscript.tex
figure ref labels added
discussion re-work

\begin{figure}
  \caption{Contact Hypothesis}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.75\textwidth]{Figures/forestBootstrappedGeneralLm.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.75\textwidth]{Figures/forestParametricFEGeneralLmer.png}
  \end{subfigure}
  \caption*{Note: \\
  (a) summary of regression results from the aggregated contact and interaction quality data.\\
  (b) summary of mixed models results of the contemporaneous contact effects.\\
  General: Fixed effects meta analytic results are presented for completeness only.}
\end{figure}

\begin{figure}
  \caption{Core Need Fulfillment}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.6\textwidth]{Figures/forestParametricFETheoryQualityCore.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.6\textwidth]{Figures/forestParametricFETheoryAttitudeCore.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.6\textwidth]{Figures/forestParametricFETheoryAttitudeCoreQuality.png}
  \end{subfigure}
  \caption*{Note: \\
  (a) Core Need Fulfillment predicting Interaction Quality.\\
  (b) Core Need Fulfillment predicting Outgroup Attitudes.\\
  (c) Core Need Fulfillment and Interaction Quality predicting Outgroup Attitudes.\\
  General: Fixed effects meta analytic results are presented for completeness only.}
\end{figure}

\begin{figure}
  \caption{Robustness Analyses}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustContact.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustSDT.png}
  \end{subfigure}
  \caption*{Note: \\
  (a) Need Fulfillment and Intergroup Contact predicting Outgroup Attitudes (full sample).\\
  (b) Core Need Fulfillment predicting Outgroup Attitudes, while controlling for self-determination theory needs (intergroup contact sample).\\
  General: Fixed effects meta analytic results are presented for completeness only.}
\end{figure}

@@

% Contact hypothesis in ESM data [inconsistent in aggregate but consistent in ML analysis]
When considering the results of the three studies jointly, we find mixed results for the basic intergroup contact hypothesis. The aggregated number of outgroup interactions across all measurements only had a significant effect on average outgroup attitudes in one of the three studies and in that study there was no significant effect of average interaction quality on outgroup attitudes (which was significant for the other two studies). This meant that only for our smallest study did we find the expected interaction between interaction frequency and interaction quality when looking at the aggregated data across all measurements. However, once we considered the effect of having an interaction with an outgroup member (vs. not having an outgroup interaction) within the participants, we find that intergroup contact is a strong predictor of outgroup attitudes across all three studies (even when taking other non-outgroup interactions into account). While this later results is in line with our expectations, the earlier aggregate findings are surprising because most past cross-sectional studies that have investigated the effect of interaction frequency have also considered an aggregate measurement (albeit not one from actual daily reports). Even though this inconsistency with past research might be a data artefact (e.g., because most people reported more substantially mode measurements during which they did not have an outgroup interaction), the absence of this aggregate effect could underline the fact that cross-sectional retrospective data might suffer from recall biases (e.g., where times with no outgroup interactions are undervalued by participants during the retrospective evaluations). In summary, we thus find that when we aggregate interaction frequencies and interaction quality from our intensive longitudinal data we find inconsistent results from the past literature but within participants intergroup contacts are significantly related with positive outgroup attitudes.

@@

% Need fulfillment in intergroup contact [consistent predictor, through quality, slightly better than Allport and SDT]
Finally, when looking at the results for our main proposal regarding the importance of situational key needs, we find highly consistent positive results across all three studies. We find that in all three extensive longitudinal datasets, the fulfillment of core needs during intergroup contacts predicts higher interaction quality perceptions and more positive outgroup attitudes. We find that in all three studies the effect of core need fulfillment on outgroup attitudes is either fully or partially through its effect on quality perceptions, and we find that need fulfillment is an important predictor even when taking other fundamental psychological needs or Allport's conditions into account. In fact we find that our core situational needs measure predicted outgroup attitudes slightly better than Allport's conditions and consistently explained more variance in outgroup attitudes than any of the other psychological needs. In most cases the core need even took over the variances previously explained by the self-determination theory needs. We thus find strong evidence that within everyday life interactions of recent migrants with majority group members, the perception that one's interaction specific psychological needs are fulfilled offers a meaningful and flexible predictor of outgroup attitudes. As such the data offers a first test of the motivational basis of intergroup contact.

>>>>

\begin{figure}
  \caption{Contact Hypothesis}
  \label{fig:ContactHypothesis}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.75\textwidth]{Figures/forestBootstrappedGeneralLm.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.75\textwidth]{Figures/forestParametricFEGeneralLmer.png}
  \end{subfigure}
  \caption*{Note: \\
  (a) summary of regression results from the aggregated contact and interaction quality data.\\
  (b) summary of mixed models results of the contemporaneous contact effects.\\
  General: Fixed effects meta analytic results are presented for completeness only.}
\end{figure}

\begin{figure}
  \caption{Core Need Fulfillment}
  \label{fig:AllportNeedFulfillment}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.6\textwidth]{Figures/forestParametricFETheoryQualityCore.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.6\textwidth]{Figures/forestParametricFETheoryAttitudeCore.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.6\textwidth]{Figures/forestParametricFETheoryAttitudeCoreQuality.png}
  \end{subfigure}
  \caption*{Note: \\
  (a) Core Need Fulfillment predicting Interaction Quality.\\
  (b) Core Need Fulfillment predicting Outgroup Attitudes.\\
  (c) Core Need Fulfillment and Interaction Quality predicting Outgroup Attitudes.\\
  General: Fixed effects meta analytic results are presented for completeness only.}
\end{figure}

\begin{figure}
  \caption{Robustness Analyses}
  \label{fig:Robustness}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustContact.png}
  \end{subfigure}
  \begin{subfigure}{\textwidth}
    \caption{}
    \centering\includegraphics[width=0.65\textwidth]{Figures/forestParametricFERobustSDT.png}
  \end{subfigure}
  \caption*{Note: \\
  (a) Need Fulfillment and Intergroup Contact predicting Outgroup Attitudes (full sample).\\
  (b) Core Need Fulfillment predicting Outgroup Attitudes, while controlling for self-determination theory needs (intergroup contact sample).\\
  General: Fixed effects meta analytic results are presented for completeness only.}
\end{figure}

@@

% Contact hypothesis in ESM data [inconsistent in aggregate but consistent in ML analysis]
When considering the results of the three studies jointly, we find mixed results for the basic intergroup contact hypothesis. For the data that aggregated people's interaction frequencies and average interaction quality we don't find a consistent predictor of outgroup attitudes. Neither interaction quality, or interaction quality, nor their interaction was a significant predictor across all studies (also see Figure \ref{fig:ContactHypothesis}a). However, once we considered the effect of having an interaction with an outgroup member (vs. not having an outgroup interaction) within the participants in a multi-level model, we find that intergroup contact is a strong predictor of outgroup attitudes across all three studies (even when taking other non-outgroup interactions into account; also see Figure \ref{fig:ContactHypothesis}b). While this later results is in line with our expectations, the earlier aggregate findings are surprising because most past cross-sectional studies that have investigated the effect of interaction frequency have also considered an aggregate measurement (albeit not one from actual daily reports). Even though this inconsistency with past research might be a data artefact (e.g., because most people reported substantially more measurements during which they did not have an outgroup interaction), the absence of this aggregate effect could underline the fact that cross-sectional retrospective data might might be misleading because it presents a mixture of within and between effects \citep[][]{Hamaker2020} or suffer from recall biases (e.g., where times with no outgroup interactions are undervalued by participants during the retrospective evaluations). In summary, we thus find that when we aggregate interaction frequencies and interaction quality from our intensive longitudinal data we find inconsistent results from the past literature but within participants intergroup contacts are significantly related with positive outgroup attitudes.

@@

% Need fulfillment in intergroup contact [consistent predictor, through quality, slightly better than Allport and SDT]
Finally, when looking at the results for our main proposal regarding the importance of situational key needs, we find highly consistent positive results across all three studies. We find that in all three extensive longitudinal datasets, the fulfillment of core needs during intergroup contacts predicts higher interaction quality perceptions and more positive outgroup attitudes. We find that in all three studies the effect of core need fulfillment on outgroup attitudes is either fully or partially through its effect on quality perceptions, and we find that need fulfillment is an important predictor even when taking other fundamental psychological needs or Allport's conditions into account. In fact we find that our core situational needs measure predicted outgroup attitudes slightly better than Allport's conditions and consistently explained more variance in outgroup attitudes than any of the other psychological needs. In most cases the core need even took over the variances previously explained by the self-determination theory needs. We thus find strong evidence that within everyday life interactions of recent migrants with majority outgroup members, the perception that one's interaction specific psychological needs are fulfilled offers a meaningful and flexible predictor of outgroup attitudes. As such the data offers a first test of the motivational basis of intergroup contact.
